---
title: "Introduction to Business Analytics with R."
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This is an assignment for the module 3 of the course Introduction to Business Analytics with R.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 

Calculate the data of today and define the format "month, day, year".

```{r}
d <- Sys.Date()
d <- format.Date(d, '%m/%d/%y')
```
Check the data type of the variable created above.
```{r}
typeof(d)
```
Install the package lubridate to use the pre defined functions of the library.
```{r}
install.packages('lubridate')
library(lubridate)
```
Create new variable for "year, month, week and day".
```{r}
d_year <- year(d)
d_year
d_month <- month(d)
d_month
d_week <- week(d)
d_week
d_day <- day(d)
d_day
```
Create new variable that is 25 days after today. 
```{r}
d_25 = as.Date(d) + 25
d_25
```
Use the function to check the difference of the date variable by days.
```{r}
difftime(d_25, d)
```
